A tool for designing digital filters for the Hankel and Fourier transforms in potential, diffusive, and wavefield modeling

### 2018-01-19 ManuscriptGeophysics2018.tex
Changes after call with Key

It is very important to note again that other scenarios might yield very
different error plots. Although the new 201\,pt filter proves to be very
accurate for these three models, it might not be the best filter in other
cases. To verify the applicability of the filter we run two different tests
over a wide range of scenarios. Figure~\ref{fig:error} shows the relative
errors of DLF using the filter \emph{Wer201} compared to QWE for a deep water
model, a shallow water model, and a land model. The model consists of an air
halfspace, a water layer in the first two examples, and a subsurface halfspace.
Source depth is 1990\,m, 10\,m, and 0.5\,m in the deep, shallow, and land case,
respectively. Receiver depth is 2000\,m, 100\,m, and 0.8\,m. The subsurface
halfspace resistivity varies from 1 to 10, 100, and 1000\,$\Omega\,$m from the
left to the right column. The error plots cover offsets from 50\,m to 20\,km on
the x-axis, and 0.01\,Hz to 10\,Hz on the y-axis. It can be seen in the result
that the filter is sufficiently precise for all practical CSEM cases, hence has
generally an error far below 1\,\% (1e-2) in all cases. The exception are large
offsets with relatively high frequencies, e.g., 10\,km offset with 8\,Hz.
%
\plot*{error}{width=\textwidth}{Relative error plots comparing the DLF
  (\emph{Wer201}) with QWE for the three cases deep water (top row), shallow
  water (middle row), and land (bottom row). The subsurface halfspace
  resistivity varies from 1 to 1000\,$\Omega\,$m from the left column to the
  right column; offsets vary from 50\,m to 20\,km, and frequencies vary from
  0.01\,Hz to 10\,Hz. The DLF provides sufficiently accurate results for all
  practical CSEM applications, only breaking down for large offsets with
  relatively high frequencies. However, the signal is usually below the noise
  level anyway in this region.}
%

@@

Figure~\ref{fig:GPR} shows the GPR example as calculated in
\cite{GEO.15.Hunziker} and \cite{GEO.17.Werthmuller}; the model parameters are
given in subplot (a). The filter used for this example for the Hankel transform
is a 2001\,pt filter, derived with the fullspace solution with $f=500\,$MHz for
the inversion and the check of quality. The whole space solution used
resistivity $200\,\Omega\,$m, relative permittivity $\varepsilon_\mr{r}=10$,
and permeability $\mu_\mr{r}=1$. The source and receiver vertical separation is
1\,m. For the Fourier transform a 4096\,pt FFT was used with regularly spaced
frequencies from 0.5\,MHz to 850\,MHz and then zero-padded up to 2048\,MHz. The
frequency result is multiplied with a Ricker wavelet with a center frequency of
250\,MHz, and a gain function ($1 + |t^3|, t$ in ns) is applied. Subplot (b)
shows the result when using the adaptive quadrature (QUAD) option of
\texttt{empymod}, which uses the QAGSE routine from the Fortran QUADPACK
library. Subplot (c) shows the result using DLF with the new 2001\,pt digital
filter. The calculation with DLF took under 9 minutes and is therefore roughly
80 times faster than the calculation with QUAD, which took roughly 11 hours and
27 minutes. Calculating the same model with QWE took 7 hours and 20 minutes.
However, QWE uses in roughly 1/3 of the calculation internally QUAD in this
example, see \cite{GEO.17.Werthmuller}. (Note that DLF was run in parallel
using 4 threads at once, taking effectively only 2 minutes and 10 seconds to
calculate.
% EMmod \citep{GEO.15.Hunziker} took for the same model about 18 hours 55
% minutes, using a 61\,pt Gauss-Kronrod integration routine.
The lagged convolution version of DLF and the splined version of QWE were used
in this comparison.)
%
\plot*{GPR}{width=\textwidth}{GPR example for the model given in (a) using
  \texttt{empymod} with (b) the adaptive quadrature (QUAD) or with (c) a
  2001\,pt digital linear filter (b). QUAD took about 11\,h 27\,min to
  calculate, whereas DLF took 2\,min 10\,s in parallel on 4 nodes, hence under
  9\,min in total.}
%

>>>>

It is very important to note again that other scenarios might yield very
different error plots. Although the new 201\,pt filter proves to be very
accurate for these three models, it might not be the best filter in other
cases. To verify the applicability of the filter we run two different tests
over a wide range of scenarios. Figure~\ref{fig:error} shows the relative
errors of DLF using the filter \emph{Wer201} compared to QWE for a deep water
model, a shallow water model, and a land model. The models consist of air
(upper halfspace), a water layer in the first two examples, and a subsurface
halfspace. Source depth is 1990\,m, 10\,m, and 0.5\,m in the deep, shallow, and
land case, respectively. Receiver depth is 2000\,m, 100\,m, and 0.8\,m. The
subsurface halfspace resistivity varies from 1 to 10, 100, and
1000\,$\Omega\,$m from the left to the right column. The error plots cover
offsets from 50\,m to 20\,km on the x-axis, and 0.01\,Hz to 10\,Hz on the
y-axis. It can be seen in the result that the filter is sufficiently precise
for all practical CSEM cases, hence has generally an error far below 1\,\%
(1e-2) in all cases. The exception are large offsets with relatively high
frequencies, e.g., 10\,km offset with 8\,Hz. In the shallow water case this
zone has a very high error for subsurface resistivities of 1\,$\Omega\,$m and
10\,$\Omega\,$m (yellow zones). However, this is due to the fact that the
amplitudes of the signal for these offsets and frequencies are in the order of
1e-25~V/m.
%
\plot*{error}{width=\textwidth}{Relative error plots comparing the DLF
  (\emph{Wer201}) with QWE for the three cases deep water (top row), shallow
  water (middle row), and land (bottom row). The subsurface halfspace
  resistivity varies from 1 to 1000\,$\Omega\,$m from the left column to the
  right column; offsets vary from 50\,m to 20\,km, and frequencies vary from
  0.01\,Hz to 10\,Hz. The DLF provides sufficiently accurate results for all
  practical CSEM applications. The exception is in the shallow water case for
  $\rho = 1\,\Omega\,$m and $10\,\Omega\,$m for large offsets and high
  frequencies (yellow zones). However, the amplitudes in these regions are in
  the order of 1e-25~V/m, and therefore far below the noise level.}
%

@@

Figure~\ref{fig:GPR} shows the GPR example as calculated in
\cite{GEO.15.Hunziker} and \cite{GEO.17.Werthmuller}; the model parameters are
given in subplot (a). The filter used for this example for the Hankel transform
is a 2001\,pt filter, derived with the fullspace solution with $f=500\,$MHz for
the inversion and the check of quality. The whole space solution used
resistivity $200\,\Omega\,$m, relative permittivity $\varepsilon_\mr{r}=10$,
and permeability $\mu_\mr{r}=1$. The source and receiver vertical separation is
1\,m. For the Fourier transform a 4096\,pt FFT was used with regularly spaced
frequencies from 0.5\,MHz to 850\,MHz and then zero-padded up to 2048\,MHz. The
frequency result is multiplied with a Ricker wavelet with a center frequency of
250\,MHz, and a gain function ($1 + |t^3|, t$ in ns) is applied. Subplot (b)
shows the result when using the adaptive quadrature (QUAD) option of
\texttt{empymod}, which uses the QAGSE routine from the Fortran QUADPACK
library. Subplot (c) shows the result using DLF with the new 2001\,pt digital
filter. The calculation with DLF took under 9 minutes and is therefore roughly
80 times faster than the calculation with QUAD, which took roughly 11 hours and
27 minutes. Calculating the same model with QWE took 7 hours and 20 minutes.
However, QWE uses in roughly 1/3 of the calculation internally QUAD in this
example, see \cite{GEO.17.Werthmuller}. Note that DLF was run in parallel using
4 threads at once, taking effectively only 2 minutes and 10 seconds to
calculate.
% EMmod \citep{GEO.15.Hunziker} took for the same model about 18 hours 55
% minutes, using a 61\,pt Gauss-Kronrod integration routine.
The lagged convolution version of DLF and the splined version of QWE were used
in this comparison.
%
\plot*{GPR}{width=\textwidth}{GPR example for the model given in (a) using
  \texttt{empymod} with (b) the adaptive quadrature (QUAD) or with (c) a
  2001\,pt digital linear filter (b). QUAD took about 11\,h 27\,min to
  calculate, whereas DLF took 2\,min 10\,s in parallel on 4 nodes, hence under
  9\,min in total.}
%
